---
title: "Life Expectancy Project"
author: "Sean Johnson"
date: "2020-12-15"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r, message=FALSE}
library(readxl)
library(janitor)
library(readr)
library(dplyr)
library(ggplot2)
library(leaps)
library(rlang)
```

## Seed
```{r}
seed = 1
set.seed(seed)
```

# Data

Data was collected from the World Heath Organization

```{r}
life_expec <- read_excel(here::here("data","life_expectancy.xlsx"))
```

# Wrangling

```{r}
life_expec <-  clean_names(life_expec)
```

*Specific names to change*

```{r}
life_expec <- life_expec %>% 
  rename(thin_10to19_years = thinness_1_19_years,
         thin_5to9_years = thinness_5_9_years,
         human_development_index = income_composition_of_resources
                          )
```

Removed the following variables as they do not add to analysis: infant_deaths, adult_mortality, and human_development_index.

```{r}
life_expec <- life_expec %>% 
  select(-infant_deaths,-adult_mortality,-human_development_index)
```

Edited specific countries from undeveloped to developed as the data was imputed incorrectly

> This caused issues with data exploration, specifically when these countries were previously labeled asundeveloped caused outliers in boxplots.

```{r}
life_expec <- life_expec%>%
  mutate(status = case_when(country=="Canada"~"Developed",
                            country=="Estonia"~"Developed",
                            country=="Finland"~"Developed",
                            country=="France"~"Developed",
                            country=="Greece"~"Developed",
                            TRUE ~ status))

```

## NA's

```{r}
sum(is.na(life_expec))
apply(X=is.na(life_expec), MARGIN = 2, FUN = sum)
```
>the largest variable with missing values is population followed by gdp.

Omitting all NA values from data to be able to run analysis

>Filling the NA values with 0 does not make sense to me as this will throw the analysis by adding weigh to the point 0.

### Adjusted dataset
```{r}
le_adj <- na.omit(life_expec)
```

*Changing categorical variable status into a numeric*

```{r}
le_adj$stat_num <- as.numeric(factor(le_adj$status)) -1
```

> Countries that are Devevloping will be represented as 1 and Developed will be 0. This will be usefull for exploratory data analysis.

### Split by status
```{r}
#ordering data set by status
le_adj <- le_adj[order(le_adj$status),]

le_developed <-le_adj%>%
 filter(status=="Developed")
le_developing <-le_adj%>%
 filter(status=="Developing")

#return to alphabetical list by country
le_adj <- le_adj[order(le_adj$country),]
```

## EDA

### Summaries
```{r}
#non vaccines or gov't
le_adj %>% 
    group_by(status) %>% 
    summarize(count = n(),
             avg_lifexp = mean(life_expectancy),
             avg_pop = mean(population),
             avg_alcohol = mean(alcohol),
             avg_hiv = mean(hiv_aids),
             avg_thinLessThan10 = mean(thin_5to9_years),
             avg_thin10plus = mean(thin_10to19_years),
             avg_bmi = mean(bmi),
             avg_under5 = mean(under_five_deaths))
#vaccines 
le_adj %>% 
    group_by(status) %>% 
    summarize(count = n(),            
             avg_hep = mean(hepatitis_b),
             avg_meas = mean(measles),
             avg_ploio= mean(polio),
             avg_dipth = mean(diphtheria))
#gov't input
le_adj %>% 
    group_by(status) %>% 
    summarize(count = n(),
             avg_totExp = mean(total_expenditure),
             avg_gdp = mean(gdp),
             avg_pctExp = mean(percentage_expenditure),
             avg_school = mean(schooling))
             
```

### Univariate eda
```{r}
list = c("life_expectancy","alcohol","hiv_aids","thin_5to9_years","thin_10to19_years",
         "bmi","under_five_deaths","hepatitis_b","measles","polio","diphtheria",
         "total_expenditure","gdp","percentage_expenditure","schooling")

for (i in list) {
  name = i
  plt<-le_adj%>%
    ggplot(aes(x = !!sym(i)))+
    geom_histogram(data = subset(le_adj, stat_num == 0),aes(fill=factor(stat_num)), binwidth = 1,alpha=0.8)+
    geom_histogram(data = subset(le_adj, stat_num == 1),aes(fill=factor(stat_num)), binwidth = 1,alpha=0.2)+
    labs(x = name, y = "Count")+
    scale_fill_manual(name="Status",values=c("#0197F6","#B2B5B6"),labels=c("Developed","Developing"))
  print(plt)
}
```


Looking at difference between avg developed and avg developing life expectancy
```{r}
ggplot(le_adj, aes(x=factor(status), y=life_expectancy, fill=status)) +
  stat_summary(fun="mean", geom="bar")+
  labs(y = "Life Expectancy", x = "Staus", title="Average global life expectancy based on status")+
    scale_fill_manual(name="Status",values=c("#0197F6","#B2B5B6"))
  
```

Further looking at differences between country status

```{r}
ggplot(data = le_adj, aes(x = life_expectancy))+
  geom_histogram(data = subset(le_adj, stat_num == 0),aes(fill=factor(stat_num)),
                 binwidth = 1, alpha = .7)+
  geom_histogram(data = subset(le_adj, stat_num == 1),aes(fill=factor(stat_num)),
                 binwidth = 1, alpha = .45)+
  geom_vline(aes(xintercept=79.4),color="blue", linetype="dashed", size=.75) +
  geom_vline(aes(xintercept=65.8),color="black", linetype="dashed", size=.75)+ labs(x = "Life Expectancy", y = "Count", title = "Life expectancy histogram based on country status")+
    scale_fill_manual(name="Status",values=c("#0197F6","#B2B5B6"),labels=c("Developed","Developing"))



```

Life expectancy with year attribute

```{r}
ggplot(data = le_adj, aes(x = year, y = life_expectancy, color= status))+
  geom_point() + 
  labs(x = "Year", y = "Life Expectancy")+
  scale_color_manual(name="Status",values=c("#0197F6","#B2B5B6"),labels=c("Developed","Developing"))

ggplot(data = le_adj, aes(x = year, y = life_expectancy, color=status))+
  geom_point() + labs(x = "Year", y = "Life Expectancy")+
  scale_color_manual(name="Status",values=c("#0197F6","#B2B5B6"),labels=c("Developed","Developing"))+
  facet_wrap(~status)

ggplot(data = le_adj, aes(x = status, y=life_expectancy))+
  geom_boxplot(data = subset(le_adj, stat_num == 0), aes(fill=factor(stat_num))) + 
  geom_boxplot(data = subset(le_adj, stat_num == 1), aes(fill=factor(stat_num))) +
  labs(x = "Year", y = "Life Expectancy", title= "Global life expectancy based on status")+
  scale_fill_manual(name="Status",values=c("#0197F6","#B2B5B6"),labels=c("Developed","Developing"))+
  facet_wrap(~year,nrow = 2,  ncol = 8,)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```
### Data for a given year
```{r}
le_new <- le_adj%>%
  filter(year==2015)%>%
  select(status,hiv_aids,thin_5to9_years,polio,diphtheria,
         bmi,gdp,percentage_expenditure,schooling)
```

*renaming for next visual*
```{r}
le_new <- le_new %>% 
    group_by(status) %>% 
    summarize(
             hiv = mean(hiv_aids),
             thin5to9 = mean(thin_5to9_years),
             polio = mean(polio),
             diphtheria = mean(diphtheria),
             bmi = mean(bmi),
             gdp = mean(gdp)/1000,
             pctExp = mean(percentage_expenditure),
             school = mean(schooling))
```

```{r}
le_new <- tidyr::pivot_longer(le_new,!status, names_to = "type",values_to = "value")
```


```{r}
ggplot(le_new, aes(type, value)) + 
  geom_bar(aes(fill = status), stat = "identity", position = "dodge")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1))+
  ggtitle("Comparison between average country status: 2015")+
  labs(x = "Parameters", y = "Average Value")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1))+
  scale_fill_manual(values = c("#0197F6","#B2B5B6"))
```

Looking at multi variable interactions

```{r}
list = c("gdp","total_expenditure","percentage_expenditure","schooling")

for (i in list) {
  name = i
  plt<-le_adj%>%
    ggplot(aes(x = life_expectancy, y = !!sym(i), color= status, size=population))+
    geom_point() + 
    labs(x = "Life Expectancy", y = name)+
    scale_color_manual(name="Status", values=c("#0197F6","#B2B5B6"))
  print(plt)
}   
```

```{r}
ggplot(data = life_expec, aes(x = year, y = life_expectancy, fill=status))+
  geom_boxplot() + labs(x = "Year", y = "Life Expectancy")+
  scale_fill_manual(values = c("#0197F6","#B2B5B6"))+
  facet_wrap(~status)

ggplot(data = le_adj, aes(x = year, y = life_expectancy, fill=status))+
  geom_boxplot() + labs(x = "year", y = "life expectancy")+
  scale_fill_manual(values = c("#0197F6","#B2B5B6"))+
  facet_wrap(~status)

```

## Correltation matrices

Note: This was after some models were tested, that linearity may be an issue and was
checked via correlation matrices

### All Countries
```{r}
corplt <- le_adj %>%
    select(life_expectancy, population, alcohol, hiv_aids, thin_5to9_years,thin_10to19_years,bmi,under_five_deaths,hepatitis_b,measles,polio,diphtheria,total_expenditure,gdp,percentage_expenditure,schooling)
cormat <- cor(corplt)
melted <- reshape::melt(cormat)
melted<-melted%>%
  rename(Var1 = X1, 
         Var2 = X2)

ggplot(data = melted, aes(x=Var1, y=Var2, fill=value)) +
  ggtitle("All Countries")+
  geom_tile(color = "white")+
 scale_fill_gradient2(low = "#B2B5B6", high = "#0197F6", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson \n Correlation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 10, hjust = 1))+
 coord_fixed()
```


### Developed
```{r}

corplot <- le_developed %>%
    select(life_expectancy, population, alcohol, hiv_aids, thin_5to9_years,thin_10to19_years,bmi,under_five_deaths,hepatitis_b,measles,polio,diphtheria,total_expenditure,gdp,percentage_expenditure,schooling)
cormat <- cor(corplt)
melted <- reshape::melt(cormat)
melted<-melted%>%
  rename(Var1 = X1, 
         Var2 = X2)
ggplot(data = melted, aes(x=Var1, y=Var2, fill=value)) +
  ggtitle("Developed Countries")+
  geom_tile(color = "white")+
 scale_fill_gradient2(low = "#B2B5B6", high = "#0197F6", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson \n Correlation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 10, hjust = 1))+
 coord_fixed()
```

### Developing
```{r}

corplot <- le_developing %>%
    select(life_expectancy, population, alcohol, hiv_aids, thin_5to9_years,thin_10to19_years,bmi,under_five_deaths,hepatitis_b,measles,polio,diphtheria,total_expenditure,gdp,percentage_expenditure,schooling)
cormat <- cor(corplt)
melted <- reshape::melt(cormat)
melted<-melted%>%
  rename(Var1 = X1, 
         Var2 = X2)

ggplot(data = melted, aes(x=Var1, y=Var2, fill=value)) + 
  ggtitle("Developing Countries")+
  geom_tile(color = "white")+
 scale_fill_gradient2(low = "#B2B5B6", high = "#0197F6", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson \n Correlation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 10, hjust = 1))+
 coord_fixed()
```
>All Countries::Life Expec High Corr: School, pct exp, gdp, dipth, polio, bmi, alcohol, tot exp, hepB
                            Low Corr: hiv, thin5to9, thin10to19, under5,measles
                            
>Not much differnet with Developed and Developing Countries

Plots of high correlations

```{r}
ggplot(data = le_adj, aes(x = thin_5to9_years, y = thin_10to19_years, color=status))+
  geom_point()+
  labs(x = "Thinness 5 to 9 years of age", y = "Thinness 10 to 19 years of age")+
  scale_color_manual(name="Status",values = c("#0197F6","#B2B5B6"))


ggplot(data = le_adj, aes(x = gdp, y = percentage_expenditure, color=status))+
  geom_point()+
  labs(x = "GDP", y = "Percentage Expenditure")+
  scale_color_manual(name="Status",values = c("#0197F6","#B2B5B6"))
```
> Use caret packages vif(). This calculates the variation inflation factors of all predictors in regression models, where high values are potentials to be dropped from the model.

# Modelin Preparation
*removing indicator and categorical variables under different handles*
```{r}
le <- le_adj[,4:19]
#develo-ped
le_ped <- le_developed[,4:19]
#develo-ping
le_ping <- le_developing[,4:19]
```

## Variable selection : mean squared error (MSE)

```{r}
set.seed(seed)
#all countries
train = le %>%
  sample_frac(0.7)
test = le %>%
  setdiff(train)

#developed countries
trained = le_ped %>%
  sample_frac(0.7)
tested = le_ped %>%
  setdiff(trained)

#developing countires
training = le_ping %>%
  sample_frac(0.7)
testing = le_ping %>%
  setdiff(training)
```

## Baseline 

```{r}
#Mean only
base_MSE = mean((mean(train$life_expectancy)-test$life_expectancy)^2)
base_MSE

base_MSE1 = mean((mean(trained$life_expectancy)-tested$life_expectancy)^2)
base_MSE1

base_MSE2 = mean((mean(training$life_expectancy)-testing$life_expectancy)^2)
base_MSE2

```

##Ordinary Least Squares - Baseline

```{r}
lm = lm(life_expectancy ~., train)
summary(lm)

test = test %>%
  mutate(predictions = predict(lm, test))

slr_MSE_test = test %>%
  summarize(slr_MSE_test = mean((life_expectancy-predictions)^2))
slr_MSE_test

lm1 = lm(life_expectancy ~., trained)
summary(lm1)

tested = tested %>%
  mutate(predictions = predict(lm1, tested))

slr_MSE_test1 = tested %>%
  summarize(slr_MSE_test1 = mean((life_expectancy-predictions)^2))
slr_MSE_test1

lm2 = lm(life_expectancy ~., training)
summary(lm2)

testing = testing %>%
  mutate(predictions = predict(lm2, testing))

slr_MSE_test2 = testing %>%
  summarize(slr_MSE_test2 = mean((life_expectancy-predictions)^2))
slr_MSE_test2

```

## Best Subsets 

### r squared
```{r}
regfit_full = regsubsets(life_expectancy ~ ., data=train) 
reg_summary = summary(regfit_full)
reg_summary
names(reg_summary)
reg_summary$rsq

regfit_full1 = regsubsets(life_expectancy ~ ., data=trained) 
reg_summary1 = summary(regfit_full1)
reg_summary1
names(reg_summary1)
reg_summary1$rsq

regfit_full2 = regsubsets(life_expectancy ~ ., data=training) 
reg_summary2 = summary(regfit_full2)
reg_summary2
names(reg_summary2)
reg_summary2$rsq

```
>developed countries data set has issues with hiv_aids

```{r}
par(mfrow = c(4,4))
plot(reg_summary$rss, type = "l")

plot(reg_summary$adjr2, xlab = "Number of Variables", ylab = "Adjusted RSq", type = "l")
adj_r2_max = which.max(reg_summary$adjr2)
points(adj_r2_max, reg_summary$adjr2[adj_r2_max], col ="red", cex = 2, pch = 20)

plot(reg_summary$cp, xlab = "Number of Variables", ylab = "Cp", type = "l")
cp_min = which.min(reg_summary$cp) 
points(cp_min, reg_summary$cp[cp_min], col = "red", cex = 2, pch = 20)

plot(reg_summary$bic, xlab = "Number of Variables", ylab = "BIC", type = "l")
bic_min = which.min(reg_summary$bic) 
points(bic_min, reg_summary$bic[bic_min], col = "red", cex = 2, pch = 20)



par(mfrow = c(2,2))
plot(reg_summary1$rss, xlab = "Number of Variables", ylab = "RSS", type = "l")

plot(reg_summary1$adjr2, xlab = "Number of Variables", ylab = "Adjusted RSq", type = "l")
adj_r2_max = which.max(reg_summary1$adjr2)
points(adj_r2_max, reg_summary1$adjr2[adj_r2_max], col ="red", cex = 2, pch = 20)

plot(reg_summary1$cp, xlab = "Number of Variables", ylab = "Cp", type = "l")
cp_min = which.min(reg_summary1$cp) 
points(cp_min, reg_summary1$cp[cp_min], col = "red", cex = 2, pch = 20)

plot(reg_summary1$bic, xlab = "Number of Variables", ylab = "BIC", type = "l")
bic_min = which.min(reg_summary1$bic) 
points(bic_min, reg_summary1$bic[bic_min], col = "red", cex = 2, pch = 20)



par(mfrow = c(2,2))
plot(reg_summary2$rss, xlab = "Number of Variables", ylab = "RSS", type = "l")

plot(reg_summary2$adjr2, xlab = "Number of Variables", ylab = "Adjusted RSq", type = "l")
adj_r2_max = which.max(reg_summary2$adjr2)
points(adj_r2_max, reg_summary2$adjr2[adj_r2_max], col ="red", cex = 2, pch = 20)

plot(reg_summary2$cp, xlab = "Number of Variables", ylab = "Cp", type = "l")
cp_min = which.min(reg_summary2$cp) 
points(cp_min, reg_summary2$cp[cp_min], col = "red", cex = 2, pch = 20)

plot(reg_summary2$bic, xlab = "Number of Variables", ylab = "BIC", type = "l")
bic_min = which.min(reg_summary2$bic) 
points(bic_min, reg_summary2$bic[bic_min], col = "red", cex = 2, pch = 20)

```

```{r}
plot(regfit_full, scale="r2")

plot(regfit_full1, scale="r2")

plot(regfit_full2, scale="r2")
```

### adj rsq
```{r}
plot(regfit_full, scale="adjr2")
coef(regfit_full, 8)

plot(regfit_full1, scale="adjr2")
coef(regfit_full1, 8)

plot(regfit_full2, scale="adjr2")
coef(regfit_full2, 8)

```

```{r}
lm = lm(life_expectancy ~ alcohol + hiv_aids + thin_5to9_years + polio + diphtheria + bmi + percentage_expenditure + schooling, train)
summary(lm)

test = test %>%
  mutate(predictions = predict(lm, test))

adj_MSE_test = test %>%
  summarize(adj_MSE_test = mean((life_expectancy-predictions)^2))
adj_MSE_test


lm1 = lm(life_expectancy ~ alcohol + hepatitis_b + under_five_deaths + total_expenditure + gdp + bmi + schooling + hiv_aids, trained)
summary(lm1)

tested = tested %>%
  mutate(predictions = predict(lm1, tested))

adj_MSE_test1 = tested %>%
  summarize(adj_MSE_test1 = mean((life_expectancy-predictions)^2))
adj_MSE_test1


lm2 = lm(life_expectancy ~ alcohol + hiv_aids + total_expenditure + polio + diphtheria + bmi + percentage_expenditure + schooling, training)
summary(lm2)

testing = testing %>%
  mutate(predictions = predict(lm2, testing))

adj_MSE_test2 = testing %>%
  summarize(adj_MSE_test2 = mean((life_expectancy-predictions)^2))
adj_MSE_test2
```

### marrows cp
```{r}
plot(regfit_full, scale="Cp")
coef(regfit_full, 8)

plot(regfit_full1, scale="Cp")
coef(regfit_full1, 8)

plot(regfit_full2, scale="Cp")
coef(regfit_full2, 8)
```

```{r}
lm = lm(life_expectancy ~ alcohol + hiv_aids + thin_5to9_years + polio + diphtheria + bmi + percentage_expenditure + schooling, train)
summary(lm)

test = test %>%
  mutate(predictions = predict(lm, test))

Cp_MSE_test = test %>%
  summarize(Cp_MSE_test = mean((life_expectancy-predictions)^2))
Cp_MSE_test


lm1 = lm(life_expectancy ~ alcohol + hepatitis_b + under_five_deaths + total_expenditure + gdp + bmi + schooling + hiv_aids, trained)
summary(lm1)

tested = tested %>%
  mutate(predictions = predict(lm1, tested))

Cp_MSE_test1 = tested %>%
  summarize(Cp_MSE_test1 = mean((life_expectancy-predictions)^2))
Cp_MSE_test1


lm2 = lm(life_expectancy ~ alcohol + hiv_aids + total_expenditure + polio + diphtheria + bmi + percentage_expenditure + schooling, training)
summary(lm2)

testing = testing %>%
  mutate(predictions = predict(lm2, testing))

Cp_MSE_test2 = testing %>%
  summarize(Cp_MSE_test2 = mean((life_expectancy-predictions)^2))
Cp_MSE_test2
```

### bic
```{r}
plot(regfit_full, scale="bic")
coef(regfit_full, 7)

plot(regfit_full1, scale="bic")
coef(regfit_full1, 7)

plot(regfit_full2, scale="bic")
coef(regfit_full2, 7)
```

```{r}
lm = lm(life_expectancy ~ alcohol + hiv_aids + polio + diphtheria + bmi + percentage_expenditure + schooling, train)
summary(lm)

test = test %>%
  mutate(predictions = predict(lm, test))

bic_MSE_test = test %>%
  summarize(bic_MSE_test = mean((life_expectancy-predictions)^2))
bic_MSE_test


lm1 = lm(life_expectancy ~ alcohol + hepatitis_b + total_expenditure + gdp + bmi + schooling + hiv_aids, trained)
summary(lm1)

tested = tested %>%
  mutate(predictions = predict(lm1, tested))

bic_MSE_test1 = tested %>%
  summarize(bic_MSE_test1 = mean((life_expectancy-predictions)^2))
bic_MSE_test1


lm2 = lm(life_expectancy ~ alcohol + hiv_aids +  polio + diphtheria + bmi + percentage_expenditure + schooling, training)
summary(lm2)

testing = testing %>%
  mutate(predictions = predict(lm2, testing))

bic_MSE_test2 = testing %>%
  summarize(bic_MSE_test2 = mean((life_expectancy-predictions)^2))
bic_MSE_test2
```

```{r}
test = test[,1:16]
tested = tested[,1:16]
testing = testing[,1:16]
```

## Best Subsets with CV

```{r}
regfit_best_train = regsubsets(life_expectancy ~ ., data=train, nvmax = 15) 
summary(regfit_best_train)

```

```{r}
test_mat = model.matrix (life_expectancy~., data = test)

```

```{r}
val_errors = rep(NA,15)

# Iterate over each size i
for(i in 1:15){
    
    # Extract the vector of predictors in the best fit model on i predictors
    coefi = coef(regfit_best_train, id = i)
    
    # Make predictions using matrix multiplication of the test matirx and the coefficients vector
    pred = test_mat[,names(coefi)]%*%coefi
    
    # Calculate the MSE
    val_errors[i] = mean((test$life_expectancy-pred)^2)
}
```


```{r}
# Find the model with the smallest error
min = which.min(val_errors)
min
# Plot the errors for each model size
plot(val_errors, type = 'b')
points(min, val_errors[min][1], col = "red", cex = 2, pch = 20)
```

```{r}
#Creating a predict function for regsubsets
predict.regsubsets = function(object,newdata,id,...){
      form = as.formula(object$call[[2]]) 
      mat = model.matrix(form,newdata)   
      coefi = coef(object,id=id)          
      xvars = names(coefi)                
      mat[,xvars]%*%coefi               
}

```


```{r}
regfit_best = regsubsets(life_expectancy~., data = train, nvmax = 15)
coef(regfit_best_train, 12)
```


```{r}
lm = lm(life_expectancy ~ alcohol + hiv_aids +thin_5to9_years + thin_10to19_years + hepatitis_b + polio + diphtheria + bmi + total_expenditure +percentage_expenditure + gdp + schooling, train)
summary(lm)

test = test %>%
  mutate(predictions = predict(lm, test))

bscv_MSE_test = test %>%
  summarize(bscv_MSE_test = mean((life_expectancy-predictions)^2))
bscv_MSE_test
```
	

```{r}
regfit_best_train1 = regsubsets(life_expectancy ~ ., data=trained, nvmax = 15) 
summary(regfit_best_train1)

```
>Note j will be 14 here



```{r}
test_mat1 = model.matrix (life_expectancy~., data = tested)

```

```{r}
val_errors1 = rep(NA,15)

# Iterate over each size j
for(j in 1:14){
    
    # Extract the vector of predictors in the best fit model on j predictors
    coefi1 = coef(regfit_best_train1, id = j)
    
    # Make predictions using matrix multiplication of the test matirx and the coefficients vector
    pred1 = test_mat1[,names(coefi1)]%*%coefi1
    
    # Calculate the MSE
    val_errors1[j] = mean((tested$life_expectancy-pred1)^2)
}
```


```{r}
# Find the model with the smallest error
min1 = which.min(val_errors1)
min1
# Plot the errors for each model size
plot(val_errors1, type = 'b')
points(min1, val_errors1[min1][1], col = "red", cex = 2, pch = 20)
```


```{r}
regfit_best1 = regsubsets(life_expectancy~., data = trained, nvmax = 15)
coef(regfit_best_train1, 12)
```


```{r}
lm1 = lm(life_expectancy ~ population + alcohol + hiv_aids  + thin_10to19_years + hepatitis_b + measles +  bmi + under_five_deaths + total_expenditure +percentage_expenditure + gdp + schooling, trained)
summary(lm1)

tested = tested %>%
  mutate(predictions = predict(lm1, tested))

bscv_MSE_test1 = tested %>%
  summarize(bscv_MSE_test1 = mean((life_expectancy-predictions)^2))
bscv_MSE_test1	
```	


```{r}
regfit_best_train2 = regsubsets(life_expectancy ~ ., data=training, nvmax = 15) 
summary(regfit_best_train2)

```

```{r}
test_mat2 = model.matrix (life_expectancy~., data = testing)

```

```{r}
val_errors2 = rep(NA,15)

# Iterate over each size k
for(k in 1:14){
    
    # Extract the vector of predictors in the best fit model on k predictors
    coefi2 = coef(regfit_best_train2, id = k)
    
    # Make predictions using matrix multiplication of the test matirx and the coefficients vector
    pred2 = test_mat2[,names(coefi2)]%*%coefi2
    
    # Calculate the MSE
    val_errors2[k] = mean((testing$life_expectancy-pred2)^2)
}
```


```{r}
# Find the model with the smallest error
min2 = which.min(val_errors2)
min2
# Plot the errors for each model size
plot(val_errors2, type = 'b')
points(min2, val_errors2[min2][1], col = "red", cex = 2, pch = 20)
```


```{r}
regfit_best2 = regsubsets(life_expectancy~., data = training, nvmax = 15)
coef(regfit_best_train2, 7)
```


```{r}
lm2 = lm(life_expectancy ~ alcohol + hiv_aids  + polio + bmi + diphtheria + percentage_expenditure + schooling, training)
summary(lm1)

testing = testing %>%
  mutate(predictions = predict(lm2, testing))

bscv_MSE_test2 = testing %>%
  summarize(bscv_MSE_test2 = mean((life_expectancy-predictions)^2))
bscv_MSE_test2	
```	

```{r}
library(glmnet)
library(pls)
```

##LASSO

```{r}
set.seed(seed)
#remove life_expectancy column
x_train = model.matrix(life_expectancy~., train)[,-1]
x_test = model.matrix(life_expectancy~., test)[,-1]
x_test = x_test[,1:15]

y_train = train %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test = test %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

grid = 10^seq(10, -2, length = 100)

lasso_mod = glmnet(x_train,y_train, alpha = 1, lambda = grid)


cv.out = cv.glmnet(x_train, y_train, alpha = 1) 
plot(cv.out)
bestlam = cv.out$lambda.min

lasso_pred = predict(lasso_mod, s = bestlam, newx = x_test) # Use best lambda to predict test data
lasso_MSE_test = mean((lasso_pred - y_test)^2) 
lasso_MSE_test

#rsq calc
rss <-sum((lasso_pred - y_test)^2)
tss <-sum((lasso_pred - mean(y_test))^2)
rsq_lasso <- 1 -rss/tss


#remove life_expectancy column
x_train1 = model.matrix(life_expectancy~., trained)[,-1]
x_test1 = model.matrix(life_expectancy~., tested)[,-1]
x_test1 = x_test1[,1:15]


y_train1 = trained %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test1 = tested %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

grid = 10^seq(10, -2, length = 100)

lasso_mod1 = cv.glmnet(x_train1, y_train1, alpha = 1, lambda=grid, thresh = 1e-12) 


cv.out1 = cv.glmnet(x_train1, y_train1, alpha = 1) 
plot(cv.out1)
bestlam1 = cv.out1$lambda.min


lasso_pred1 = predict(lasso_mod1, s = bestlam1, newx = x_test1)
lasso_MSE_test1 = mean((lasso_pred1 - y_test1)^2)
lasso_MSE_test1




#remove life_expectancy column
x_train2 = model.matrix(life_expectancy~., training)[,-1]
x_test2 = model.matrix(life_expectancy~., testing)[,-1]
x_test2 = x_test2[,1:15]


y_train2 = training %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test2 = testing %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

grid = 10^seq(10, -2, length = 100)

lasso_mod2 = cv.glmnet(x_train2, y_train2, alpha = 1, lambda=grid, thresh = 1e-12) 


cv.out2 = cv.glmnet(x_train2, y_train2, alpha = 1) 
plot(cv.out2)
bestlam2 = cv.out2$lambda.min


lasso_pred2 = predict(lasso_mod2, s = bestlam2, newx = x_test2)
lasso_MSE_test2 = mean((lasso_pred2 - y_test2)^2)
lasso_MSE_test2

```
# Extra
## PCR

```{r}
set.seed(seed)

pcr_fit = pcr(life_expectancy~., data = train, scale = TRUE, validation = "CV")
summary(pcr_fit)

#have to remove variable hiv_aids as it is causing an infite loop
pcr_fit1 = pcr(life_expectancy~.-hiv_aids, data = trained, scale = TRUE, validation = "CV")
summary(pcr_fit1)


pcr_fit2 = pcr(life_expectancy~., data = training, scale = TRUE, validation = "CV")
summary(pcr_fit2)

```
 
*Finding number of components*

```{r}
validationplot(pcr_fit, val.type = "MSEP")
validationplot(pcr_fit1, val.type = "MSEP")
validationplot(pcr_fit2, val.type = "MSEP")

```


```{r}
x_train = model.matrix(life_expectancy~., train)[,-1]
x_test = model.matrix(life_expectancy~., test)[,-1]
x_test = x_test[,1:15]

y_train = train %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test = test %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

pcr_pred = predict(pcr_fit, x_test, ncomp=13)
pcr_MSE_test = mean((pcr_pred-y_test)^2)
pcr_MSE_test



x_train1 = model.matrix(life_expectancy~.-hiv_aids, trained)[,-1]
x_test1 = model.matrix(life_expectancy~.-hiv_aids, tested)[,-1]
x_test1 = x_test1[,1:14]

y_train1 = trained %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test1 = tested %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

pcr_pred1 = predict(pcr_fit1, x_test1, ncomp=14)
pcr_MSE_test1 = mean((pcr_pred1-y_test1)^2)
pcr_MSE_test1



x_train2 = model.matrix(life_expectancy~., training)[,-1]
x_test2 = model.matrix(life_expectancy~., testing)[,-1]
x_test2 = x_test2[,1:15]

y_train2 = training %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test2 = testing %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

pcr_pred2 = predict(pcr_fit2, x_test2, ncomp=15)
pcr_MSE_test2 = mean((pcr_pred2-y_test2)^2)
pcr_MSE_test2

```

##PLS

```{r}
set.seed(seed)

pls_fit = plsr(life_expectancy~., data = train, scale = TRUE, validation = "CV")
summary(pls_fit)

#removed hiv_aids variable as it was causing a loop error
pls_fit1 = plsr(life_expectancy~.-hiv_aids, data = trained, scale = TRUE, validation = "CV")
summary(pls_fit1)


pls_fit2 = plsr(life_expectancy~., data = training, scale = TRUE, validation = "CV")
summary(pls_fit2)

```



```{r}
validationplot(pls_fit, val.type = "MSEP")

validationplot(pls_fit1, val.type = "MSEP")

validationplot(pls_fit2, val.type = "MSEP")
```


```{r}
x_train = model.matrix(life_expectancy~., train)[,-1]
x_test = model.matrix(life_expectancy~., test)[,-1]
x_test = x_test[,1:15]

y_train = train %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test = test %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

pls_pred = predict(pls_fit, x_test, ncomp = 8)
pls_MSE_test = mean((pls_pred - y_test)^2)
pls_MSE_test



x_train1 = model.matrix(life_expectancy~.-hiv_aids, trained)[,-1]
x_test1 = model.matrix(life_expectancy~.-hiv_aids, tested)[,-1]
x_test1 = x_test1[,1:14]

y_train1 = trained %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test1 = tested %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

pls_pred1 = predict(pls_fit1, x_test1, ncomp = 5)
pls_MSE_test1 = mean((pls_pred1 - y_test1)^2)
pls_MSE_test1




x_train2 = model.matrix(life_expectancy~., training)[,-1]
x_test2 = model.matrix(life_expectancy~., testing)[,-1]
x_test2 = x_test2[,1:15]

y_train2 = training %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

y_test2 = testing %>%
  select(life_expectancy) %>%
  unlist() %>%
  as.numeric()

pls_pred2 = predict(pls_fit2, x_test2, ncomp = 9)
pls_MSE_test2 = mean((pls_pred2 - y_test2)^2)
pls_MSE_test2
```
Load caret, see if there is any variance inflation
```{r}
library(caret)
```


# All Countries
>The lowest MSE average from all multiple seeds was found using Lasso Regression method.

```{r}
coef(cv.out)
rsq_lasso
plot(cv.out)


lm = lm(life_expectancy ~population+alcohol+hiv_aids+thin_5to9_years+polio+diphtheria+bmi+gdp+percentage_expenditure+schooling, le_adj)
modsum =summary(lm)
r2 = modsum$r.squared
r2 = round(r2, digits=4)
r2
rsq = "Rsq = " 
ggplot(le_adj,aes(x=year,y=life_expectancy))+
  geom_point()+
  labs(title= "All countries", y="Life Expectancy", x = "Year")+
  geom_smooth(se=FALSE, method="lm")+
  #geom_label(label=r2, x=2012, y=40,)+
  geom_label(label=paste(rsq,r2), x=2012, y=40)
```
>best model is life_expectancy~ hiv_aids+thin_5to9_years+polio+diphtheria+bmi+gdp+percentage_expenditure+schooling
>mse is 19.67224


#All Developed
>The lowest MSE average from all 10 seeds was found using the OLS method.

```{r}
#removed hiv_aids due to errors

lm1 = lm(life_expectancy ~.-hiv_aids, trained)
slr_MSE_test1

car::vif(lm1)

summary(lm1)
```

```{r}
#removed hiv_aids due to errors

lm1 = lm(life_expectancy ~.-hiv_aids-thin_5to9_years, trained)
tested = tested %>%
  mutate(predictions = predict(lm1, tested))

slr_MSE_test1 = tested %>%
  summarize(slr_MSE_test1 = mean((life_expectancy-predictions)^2))
slr_MSE_test1

car::vif(lm1)

summary(lm1)
```




```{r}

#removed hiv_aids due to run error.

lm1 = lm(life_expectancy ~.-hiv_aids-thin_5to9_years-percentage_expenditure-total_expenditure, trained)
tested = tested %>%
  mutate(predictions = predict(lm1, tested))

slr_MSE_test1 = tested %>%
  summarize(slr_MSE_test1 = mean((life_expectancy-predictions)^2))
slr_MSE_test1

car::vif(lm1)

summary(lm1)
```

```{r}

#removed hiv_aids due to errors

lm1 = lm(life_expectancy ~.-hiv_aids-thin_5to9_years-percentage_expenditure-total_expenditure-polio, trained)
tested = tested %>%
  mutate(predictions = predict(lm1, tested))

slr_MSE_test1 = tested %>%
  summarize(slr_MSE_test1 = mean((life_expectancy-predictions)^2))
slr_MSE_test1

car::vif(lm1)

summary(lm1)
plot(lm1)


lm = lm(life_expectancy ~population+alcohol+thin_10to19_years+hepatitis_b+measles+polio+diphtheria+bmi+under_five_deaths+gdp+schooling+thin_5to9_years+percentage_expenditure+total_expenditure, le_developed)
modsum =summary(lm)
r2 = modsum$r.squared
r2 = round(r2, digits=4)
r2
rsq = "Rsq = " 
ggplot(le_developed,aes(x=year,y=life_expectancy))+
  geom_point()+
  labs(title= "Developed countries", y="Life Expectancy", x = "Year")+
  geom_smooth(se=FALSE, method="lm")+
  #geom_label(label=r2, x=2012, y=40,)+
  geom_label(label=paste(rsq,r2), x=2012, y=70)
```



>Best variable selection is life_expectancy ~ population + alcohol + thin_10to19_years +hepatitis_b+measles+diphtheria+bmi+under_five_deaths+gdp_schooling
if there is absolute colinearity.

>Otherwise the variuable selection is life_expectancy~population+alcohol+thin_10to19_years+hepatitis_b+measles+polio+diphtheria+bmi+under_five_deaths+gdp+schooling+thin_5to9_years+percentage_expenditure+total_expenditure



#All Developing

>The lowest MSE average from all 10 seeds was found using the Best Subset with Cross-Validation method.


```{r}
lm2 = lm(life_expectancy ~ alcohol + hiv_aids  + polio + bmi + diphtheria + percentage_expenditure + schooling, training)

car::vif(lm2)
bscv_MSE_test2	

plot(lm2)


lm = lm(life_expectancy ~ alcohol+hiv_aids+polio+bmi+diphtheria+percentage_expenditure+schooling, le_developing)
modsum =summary(lm)
r2 = modsum$r.squared
r2 = round(r2, digits=4)
r2
rsq = "Rsq = " 
ggplot(le_developing,aes(x=year,y=life_expectancy))+
  geom_point()+
  labs(title= "Developing countries", y="Life Expectancy", x = "Year")+
  geom_smooth(se=FALSE, method="lm")+
  #geom_label(label=r2, x=2012, y=40,)+
  geom_label(label=paste(rsq,r2), x=2012, y=45)
```

> all vif for variables are low

>variable selection for developing countries is:
life_expectancy ~ alcohol + hiv_aids  + polio + bmi + diphtheria + percentage_expenditure + schooling


.